Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/concrete-relational-model.tex
revisions to the relational motivation section
revise sections on relational model

\section{Constructing a Concrete Model}\label{sec:concrete-model}

In this section, we complete the construction of a relational model of gradual
typing. To manage the complexity, we carry out the construction in phases,
beginning with a relational model consisting of partially-ordered sets with a
notion of bisimilarity and then adding requirements for a structured set of
delays on each type as well as a notion of representability for relations that
models the cast rules in the intensional setting.

@@

In the previous section, we introduced the notion of weak bisimilarity as a
relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
parameterized by a relation on $A$. Thus, in order to define a compositional
model for all types, we need to equip the objects (i.e. the posets and error
domains) with a reflexive, symmetric bisimilarity relation. This leads to our
first new definition:

@@

Next we define a notion of \emph{error domain relation} between error domains $B$ and
$B'$. An error domain relation $d : B \rel B'$ is a relation on the underlying
posets that preserves error and respects $\theta$. More formally:
%
\begin{definition}
  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
  \end{enumerate}
\end{definition}
% Likewise, the definition of error domain relations is unchanged.

The definitions of squares for predomains are also unchanged from the previous
section, except that the morphisms on the left and right, are now morphisms of
predomains. We likewise define a notion of \textbf{error domain square}: Let
$d_i : B_i \rel B_i'$ and $d_o : B_o \rel B_o'$ be error domain relations, and
let $\phi : B_i \multimap B_o$ and $\phi' : B_i' \multimap B_o'$ be morphisms of
error domains. We say that there is an error domain square $\phi \ltsq{d_i}{d_o}
\phi'$ exactly when there is a predomain square involving the underlying
predomain morphisms and relations.

\begin{comment}
We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on
the underlying sets that is downward-closed under $\le_A$ and upward-closed
under $\le_{A'}$. More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
is a relation $R$ between the underlying sets such that:
\begin{enumerate}
    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
\end{enumerate}
\end{definition}
%
Composition of relations on predomains is the usual relational composition
(which is truncated to be propositional). Similarly, we define a (monotone)
relation on error domains to be a relation on the underlying predomains that
respects error and preserves $\theta$:
%
\begin{definition}
    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
    \begin{enumerate}
       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
    \end{enumerate}
\end{definition}
%
Because of a technical restriction involving the interaction between
propositional truncation and the later modality, we cannot define the
composition of error domain relations as simply the composition of their
underlying predomain relations. Instead, we define composition of error domain
relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
%
\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
    {b_1 \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = DnClosed]
    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
    {b_1' \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = UpClosed]
    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
    {b_1 \mathbin{d \relcomp d'} b_3'}

    \inferrule*[right = PresErr]
    { }
    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = PresTheta]
    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
\end{mathpar}
%
% We note that this composition has the following universal property.
%

We now describe the squares. Suppose we are given predomains $A_i, A_o, A_i'$,
and $A_o'$, relations $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$, and
morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with these
morphisms and relations, we say that the square commutes, written $f \le f'$, if
for all $x \in A_i$ and $x' \in A_i'$ with $(x, x') \in c_i$, we have $(f(x),
f'(x')) \in c_o$. We make the analogous definition for error domains.

\end{comment}

@@

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
extend to predomains, error domains, and their relations. The functor $U$ simply
takes the underlying poset relation of an error domain relation. Given a
relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
\times c_2 : (A_1 \times A_2) \rel (A_1' \times A_2')$ is defined by $((x, y),
(x', y')) \in (c_1 \times c_2)$ if and only if $(x, x') \in c_1$ and $(y, y')
\in c_2$. Given a relation $c : A \rel A'$ and an error domain relation $d : B
\rel B'$, the relation $c \arr d : (A \arr B) \rel (A' \arr B')$ is the error
domain relation defined by $(f, g) \in (c \arr d)$ if and only if for all $x :
A$ and $x' : A'$ with $(x , x') \in c$ we have $(f(x), g(x')) \in Ud$. The fact
that this relation preserves error and respects $\theta$ follows from the
corresponding properties for $d$.

% Free error domain
The functor $\li$ takes a predomain $A$ to the free error domain on $A$. We first
define the predomain $U\li A$. The underlying set of $U\li A$ is defined to be $\li
|A|$ (the guarded lift monad applied to $|A|$). The ordering relation is the
lock-step error ordering introduced in the previous section (taking $A = A'$ in
the definition), and the bisimilarity relation is the weak bisimilarity relation
on $\li |A|$ also defined in the previous section. With these relations, the
constructors $\eta$ and $\theta$ of the guarded lift monad are in fact morphisms
of predomains, i.e., they are monotone and preserve bisimilarity.

We observe that this predomain structure extends to an error domain structure by
noting that the required error element is given by the constructor $\mho$ and
the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
that the delay morphism $\delta = \theta \circ \nxt$ satisfies $x \bisim
\delta\, x$ for all $x$ as is required by the definition of error domain.

The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
to an error domain morphism $\li A \to B$; with this we define the action of $\li$ on
morphisms: Given $f : A \to A'$ we observe that $\eta \circ f : A \to U\li A'$, so
we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.

Finally, we define the action of $\li$ on a relation $c$ to be the heterogeneous
lock-step error ordering between $U\li A$ and $U\li A'$ and note that it by
definition satisfies the additional requirements of being an error domain
relation between the error domains $\li A$ and $\li A'$.

% The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
% a predomain $A$, we define the error domain $FA$ whose underlying predomain is
% $\li A$, whose error element is $\mho$, and whose $\theta$ map is the
% constructor $\theta$.

% It is easily verified that $\li A$ is the free error- and later-algebra on the
% predomain $A$, so we have that $\li$ is left-adjoint to $U$.

% i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
% predomain morphisms from $A$ to $UB$.

% We define $\delta : U(\li A) \To U(\li A)$ by $\delta(x) = \theta(\nxt x)$.
% We define $\hat{\delta} : \li A \arr \li A$ by $\hat{\delta} = \ext{(\delta \circ \eta)}{}$.
% Note that by definition of $\text{ext}$, we have that $\hat{\delta}$ is a morphism of error domains.

%\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}

% We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
% ``heterogeneous" version of the lock-step error ordering.

% TODO action of \li on commuting squares

\begin{comment}

% internal hom for predomains and error domains
Given predomains $A$ and $A'$, we can form the predomain of
predomain morphisms from $A$ to $A'$, denoted $A \To A'$.
\begin{itemize}
    % Should we give the definition involving x and x'?
    \item The ordering is defined by $f \le_{A \To A'} f'$ iff for all
    $x \in A$, we have $f(x) \le_{A'} f'(x)$.
    \item The bisimilarity relation is defined by $f \bisim_{A \To A'} f'$ iff
    for all $x, x' \in A$ with $x \bisim_{A} x'$, we have $f(x) \bisim_{A'} f'(x')$. 
\end{itemize}
%
Given $f : A_1' \to A_1$ and $g : A_2 \to A_2'$ we define the predomain morphism
$f \To g : (A_1 \To A_2) \to (A_1' \To A_2')$ by $\lambda h. \lambda x'. g(h(f(x')))$.

% TODO: include this?
% The monadic extension operation $\ext{\cdot}{} : (A_1 \To U (\li A_2)) \To (\li A_1 \To U(\li A_2))$
% is a morphism of predomains from $A_1 \To U(\li A_2)$ to $U(\li A_1) \To U(\li A_2)$, i.e.,
% it preserves the ordering and bisimilarity relations.

% Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
We note that $A \To UB$ carries a natural error domain structure
(in the below, the lambda is a meta-theoretic notation):
\begin{itemize}
    \item The error is given by $\lambda x . \mho_B$
    \item The $\theta$ operation is defined by
      \[ \theta_{A \To UB}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
\end{itemize}

Given a predomain $A$ and error domain $B$, we define
$A \arr B$ to be the error domain such that $U(A \arr B) = A \To UB$,
and whose error and $\theta$ operations are as defined above.
We can define the functorial action of $\arr$ on morphisms
$f \arr \phi$ in the obvious way.
%
It is easily verified that $A \arr B$ is an exponential of $UB$ by $A$
in the category of predomains and their morphisms.

\end{comment}

\begin{comment}
Lastly, given a relation of predomains $c$ between $A$ and $A'$, and a relation
of error domains $d$ between $B$ and $B'$, we define the relation $c \arr d$
between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
is related to $g \in A' \arr B'$ iff for all $x \in A$ and $x' \in A'$ with
$(x, x') \in c$, we have $(f(x), g(x')) \in Ud$.
%
One can verify that this relation is indeed a relation of error domains
in that it respects error and preserves $\theta$.
\end{comment}

@@

We now introduce the additional definitions needed to complete the construction
of our model. Recall from Section \ref{sec:adjusting-cast-rules} that in order
to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
rule by inserting a ``delay'' on the right-hand side. A similar adjustment is
needed for the DnR rule for $\iarr$. Moreover, the need to insert delays impacts
the semantics of the cast rules for \emph{all} relations, because of the
functorial nature of casts. That is, the upcast at a function type $c_i \ra c_o$
involves a downcast in the domain and an upcast in the codomain. This has two
consequences: first, the squares corresponding to the rules UpL and UpR may also
require the insertion of a delay. Second, we need to be able to insert
``higher-order'' delays in a way that follows the structures of the casts.

To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
perturbations as actual endomorphisms on $A$. Moreover, we want the resulting
endomorphisms to be extensionally equivalent to the identity function, a
condition we express by saying that they are weakly bisimilar to the identity
morphism on $A$. More formally, for every $A$ we require a homomorphism of
monoids $i_A : M_A \to \morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$.
We likewise equip every error domain $B$ with a monoid $M_B$ and interpretation
homomorphism $i_B : M_B \to \morbisimid{B}$. These concepts deserve their own
definitions: We call the triple $(A, M_A, iA)$ a \textbf{value object} as it
will be the denotation of value types in our final model. Likewise, the triple
$(B, M_B, i_B)$ will be called a \textbf{computation object}.

We must now impose an additional condition on predomain relations $c$ and error
domain relations $d$ that specify how they interact with these perturbations. In
particular, we need to be able to ``push'' and ``pull'' perturbations along
relations $c$ and $d$. The intuition for this requirement comes from the
construction of the square corresponding to the original UpL rule from the
square for the simplified version as was shown in Section
\ref{sec:towards-relational-model}. Recall that we horizontally composed the
simplified UpL square for $c$ with the identity square for $c'$ on the right.
Now that the square for UpL involves a perturbation on the right instead of the
identity, this construction will not work unless we can ``push'' the
perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
and a perturbation $m_A \in M_A$, we need to be able to turn it into a
perturbation $m_{A'} \in M_{A'}$ such that the resulting endomorphisms obtained
by applying the respective homomorphisms $i_A$ and $i_{A'}$ to the perturbations
form a square. Dually, we will need to ``pull'' a perturbation from right to
left. This notion is made formal by the following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c
    : A \rel A'$ be a predomain relation. A \textbf{push-pull structure} for $c$
    consists of monoid homomorphisms $\push : M_A \to M_A'$ and $\pull : M_A'
    \to M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A)
    \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a
    square $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
\end{definition}
%
We make the analogous definition for computation objects.

With the notion of perturbation defined, we can now discuss the final aspect of
the model, namely the representability of relations.
%
\begin{definition}[quasi-left-representable relations]
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) UpL: $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
(2) UpR: $i_A(\delle_c) \ltsq{r(A)}{c} e$.
\end{definition}
%
Observe that this generalizes the previous notion of representability, since
under that definition the perturbations were always the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.
%
Likewise, we define quasi-right-representability as follows:
%
\begin{definition}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) DnL: $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
(2) DnR: $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

Although it makes sense to refer to the quasi-representability of an arbitrary
predomain or error domain relation, quasi-representable relations do not in
general compose. In order to construct the needed squares for the composition of
two relations, we require that the relations have push-pull structures.

@@

The reason for requiring $\li c$ to be quasi-right-representable in the definition
of value relations is so that we can define the action of the functor $\li$ on
relations. Similarly, we require $Ud$ to be quasi-left-representable in the
definition of computation relations so that we can define the action of $U$ on
relations.

@@

We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
i_A)$ we define $M_{\li A}$ to be $\mathbb{N} \oplus M_A$, where $\oplus$
denotes the free product of monoids (the coproduct in the category of monoids).
The intuition behind needing to include $\mathbb{N}$ comes from the example of
the downcast for $\iarr$. Recall that we adjusted the squares corresponding to
the DnL and DnR rules by adding $(\delta \circ \eta)^\dagger$ on the side
opposite the downcast. Thus, we need a syntactic perturbation in $M_{\li A}$
that will be interpreted as $(\delta \circ \eta)^\dagger$. We accomplish this by
taking the coproduct with $\mathbb{N}$.
%
Then to interpret the perturbations as endomorphisms, by the universal property
of the coproduct of monoids it suffices to define two homomorphisms, one from
$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
endomorphisms are bisimilar to the identity since the action of $\li$ on
morphisms preserves bisimilarity.

@@

\eric{TODO: Quasi-equivalence of the relations mentioned in the syntax section.}

@@

Now we define the perturbations for $D$. Recall that to each predomain $A$ we
associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
endomorphisms bisimilar to the identity, and likewise for error domains. It may
seem as though we need to define the perturbations for $D$ by guarded recursion,
but in fact we define them via least-fixpoint in the category of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
% Unfolding the definitions, this is the same as
% %
% \( M_D \cong (M_D \times M_D) \oplus (\mathbb{N} \times M_D^{op} \times \mathbb{N} \times M_D). \)
% %
We now explain how to interpret these perturbations as endomorphisms.

@@

We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we must
insert a delay on the side opposite the downcast. In terms of perturbations,
this means that we take the perturbations $\dellp$ and $\delrp$ in the
definition of quasi-right-representability to both be $\inl\, 1$, where $\inl$
is the injection into the coproduct of monoids. We recall that the monoid of
perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and that the
interpretation homomorphism maps the generator $1$ to the endomorphism of error
domains $(\delta \circ \eta)^\dagger$.

With this choice for the perturbations, it is straightforward to show that the
DnL and DnR squares exist using the definition of the downcast and the
definition of the relation $\iarr$.

% We define $\dellp_D = \delrp_D = \delta_{FD}$. Then it is easy to show using the
% definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
%
It is also straightforward to establish the retraction property for each of
these three relations. In the case of $\iarr$, we have that the property holds
up to a delay.

% \subsection{Obtaining an Extensional Model}

\subsection{Defining an Extensional Model}

>>>>

\section{A Guarded Model of Graduality}\label{sec:concrete-relational-model}

In this section, we describe in detail our model of gradual typing. To manage
the complexity, we carry out the construction in two phases, beginning with a
relational model consisting of partially-ordered sets with a notion of
bisimilarity and then adding requirements for a structured set of syntactic
perturbations on each type as well as the notion of quasi-representability for
relations that models the cast rules. We then discuss the construction of the
dynamic type and the refined denotation of term precision that relates terms
that can be synchronized to be in lock-step. We conclude by demonstrating the
adequacy of our model with respect to the graduality property.

@@

In the previous section, we introduced the notion of weak bisimilarity as a
relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
parameterized by a relation on $A$. To define a compositional model for all
types, we equip the posets and error domains with a reflexive, symmetric
bisimilarity relation. This leads to our first new definition:

@@

Next we define an analogous notion of \emph{relation} between error domains $B$
and $B'$. An error domain relation $d : B \rel B'$ is a relation on the
underlying posets that preserves error and respects $\theta$. More formally:
%
\begin{definition}
  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
  \end{enumerate}
\end{definition}
%
As with predomain relations, we can compose error domain relations. However,
because of a technical restriction involving the interaction between
propositional truncation and the later modality, we cannot define the
composition of error domain relations as the composition of their underlying
predomain relations. Instead, we define composition of error domain relations
$d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ inductively to be the least
relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
% %
% \begin{mathpar}
%     \inferrule*[right = Comp]
%     {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
%     {b_1 \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = DnClosed]
%     {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
%     {b_1' \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = UpClosed]
%     {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
%     {b_1 \mathbin{d \relcomp d'} b_3'}

%     \inferrule*[right = PresErr]
%     { }
%     {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = PresTheta]
%     {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
%     {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
% \end{mathpar}
% %

The definitions of squares for predomains are also unchanged from the previous
section, except that the morphisms on the left and right are now morphisms of
predomains. We likewise define a notion of \textbf{error domain square}: Let
$d_i : B_i \rel B_i'$ and $d_o : B_o \rel B_o'$ be error domain relations, and
let $\phi : B_i \multimap B_o$ and $\phi' : B_i' \multimap B_o'$ be morphisms of
error domains. We say that there is an error domain square $\phi \ltsq{d_i}{d_o}
\phi'$ exactly when there is a predomain square involving the underlying
predomain morphisms and relations.

@@

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
extend to predomains, error domains, and their relations. The functor $U$ simply
takes the underlying predomain relation of an error domain relation. Given a
relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
\times c_2 : (A_1 \times A_2) \rel (A_1' \times A_2')$ is defined by $((x, y),
(x', y')) \in (c_1 \times c_2)$ if and only if $(x, x') \in c_1$ and $(y, y')
\in c_2$. Given a relation $c : A \rel A'$ and an error domain relation $d : B
\rel B'$, the relation $c \arr d : (A \arr B) \rel (A' \arr B')$ is the error
domain relation defined by $(f, g) \in (c \arr d)$ if and only if for all $x :
A$ and $x' : A'$ with $(x , x') \in c$ we have $(f(x), g(x')) \in Ud$. The fact
that this relation preserves error and respects $\theta$ follows from the
corresponding properties for $d$.

% Free error domain
The functor $\li$ takes a predomain $A$ to the free error domain on $A$. We first
define the predomain $U\li A$. The underlying set of $U\li A$ is defined to be $\li
|A|$ (the guarded lift monad applied to $|A|$). The ordering relation is the
lock-step error ordering introduced in the previous section (taking $A = A'$ in
the definition), and the bisimilarity relation is the weak bisimilarity relation
on $\li |A|$ also defined in the previous section. With these relations, the
constructors $\eta$ and $\theta$ of the guarded lift monad are in fact morphisms
of predomains, i.e., they are monotone and preserve bisimilarity.
%
We observe that this predomain structure extends to an error domain structure by
noting that the required error element is given by the constructor $\mho$ and
the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
that the delay morphism $\delta = \theta \circ \nxt : UB \to UB$ satisfies $x \bisim
\delta\, x$ for all $x$ as is required by the definition of error domain.

The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
to an error domain morphism $\li A \multimap B$; with this we define the action of $\li$ on
morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.

Finally, we define the action of $\li$ on a predomain relation $c$ to be the
heterogeneous lock-step error ordering between $U\li A$ and $U\li A'$ and note
that it by definition satisfies the additional requirements of being an error
domain relation between the error domains $\li A$ and $\li A'$.

@@

We now introduce the additional definitions needed to complete the construction
of our model. Recall from Section \ref{sec:lock-step-and-weak-bisim} that in order
to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
rule by inserting a ``delay'' on the right-hand side. A similar adjustment is
needed for the DnR rule for $\iarr$. Moreover, the need to insert delays impacts
the semantics of the cast rules for \emph{all} relations, because of the
functorial nature of casts. That is, the upcast at a function type $c_i \ra c_o$
involves a downcast in the domain and an upcast in the codomain. This has two
consequences: first, the squares corresponding to the rules UpL and UpR may also
require the insertion of a delay. Second, we need to be able to insert
``higher-order'' delays in a way that follows the structures of the casts.

To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
syntactic perturbations as \emph{semantic} perturbations on $A$, i.e., as
endomorphisms that are bisimilar to the identity.
% Moreover, we want the
% resulting endomorphisms to be extensionally equivalent to the identity function,
% a condition we express by saying that they are weakly bisimilar to the identity
% morphism on $A$. 
More formally, for every $A$ we require a homomorphism of monoids $i_A : M_A \to
\morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$. We likewise equip
every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
: M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
denotation of value types in our final model. Likewise, the triple $(B, M_B,
i_B)$ will be called a \textbf{computation object}.

We must impose an additional condition on predomain relations $c$ and error
domain relations $d$ that specify how they interact with these perturbations. In
particular, we need to be able to ``push'' and ``pull'' perturbations along
relations $c$ and $d$. The intuition for this requirement comes from the
construction of the square corresponding to the original UpL rule from the
square for the simplified version as was shown in Section
\ref{sec:towards-relational-model}. Recall that we horizontally composed the
simplified UpL square for $c$ with the identity square for $c'$ on the right.
Now that the square for UpL involves a perturbation on the right instead of the
identity, this construction will not work unless we can ``push'' the
perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
and a syntactic perturbation $m_A \in M_A$, we need to be able to turn it into a
syntactic perturbation $m_{A'} \in M_{A'}$ such that the resulting semantic
perturbations obtained by applying the respective homomorphisms $i_A$ and
$i_{A'}$ form a square. Dually, we will need to ``pull'' a perturbation from
right to left. This notion is made formal by the following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c
    : A \rel A'$ be a predomain relation. A \textbf{push-pull structure} for $c$
    (with respect to $M_A$ and $M_{A'}$) consists of monoid homomorphisms $\push
    : M_A \to M_A'$ and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$
    there is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'}) \ltsq{c}{c}
    i_{A'}(m_{A'})$.
\end{definition}
%
We make the analogous definition for computation objects.

With the notion of syntactic perturbation defined, we can now make formal the
final aspect of the model: the notion of quasi-represntability of relations.
%
\begin{definition}[quasi-left-representable relations]
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
\end{definition}
%
Observe that this weakens the previous notion of representability, since under
that definition the perturbations were required to be the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.

Likewise, we define quasi-right-representability as follows:
%
\begin{definition}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \multimap B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) $\dnl$: $p \ltsq{\le_{B'}}{d} i_{B'}(\delrp_d)$ and
(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

Given $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$, it makes sense to refer to the
quasi-representability of a predomain relation $c : A \rel A'$. However,
quasi-representable relations do not in general compose: in order to construct
the needed squares for the composition of two relations, we require that the
relations have push-pull structures.

@@

The reason for requiring $\li c$ to be quasi-right-representable in the
definition of value relations is so that we can define the action of the functor
$\li$ on relations, taking a value relation to a computation relation. Likewise,
we require $Ud$ to be quasi-left-representable in the definition of computation
relations so that we can define the action of $U$ on relations.

@@

We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
from the example of the downcast for $\iarr$. Recall that we adjusted the
squares corresponding to the DnL and DnR rules by adding $(\delta \circ
\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
%
Then to interpret the perturbations as endomorphisms, by the universal property
of the coproduct of monoids it suffices to define two homomorphisms, one from
$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
endomorphisms are bisimilar to the identity since the action of $\li$ on
morphisms preserves bisimilarity.

@@

@@

Now we define the perturbations for $D$. Recall that to each predomain $A$ we
associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
endomorphisms bisimilar to the identity, and likewise for error domains. It may
seem as though we need to define the perturbations for $D$ by guarded recursion,
but in fact we define them via least-fixpoint in the category of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
We now explain how to interpret these perturbations as endomorphisms.

@@

We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we
must insert a delay on the side opposite the downcast. In terms of syntactic
perturbations, this means that we take $\dellp$ and $\delrp$ in the definition
of quasi-right-representability to both be $\inl\, 1$, where $\inl$ is the
injection into the coproduct of monoids. We recall that the definition of
syntactic perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and
that the interpretation homomorphism maps the generator $1$ to the endomorphism
of error domains $(\delta \circ \eta)^\dagger$.
%
With this choice for the perturbations, it is straightforward to show that the
DnL and DnR squares exist using the definition of the downcast and the
definition of the relation $\iarr$.

It is also straightforward to establish the retraction property for each of
these three relations. In the case of $\iarr$, we have that the property holds
up to a delay.

\subsection{Interpreting Term Precision: Extensional Squares}
